Approximation as Differentiation

### 2025-05-24 main/category-of-families.tex
Rename file.

The goal now is to equip the interpretation of a (first-order) program with a Galois connection relating
approximations of its inputs to approximations of its output. Since the specific lattices of approximations
depend on the value being approximated, the semantics of every type $\tau$ must have both a ``conventional''
component, providing the set $X$ which interprets $\tau$, plus an associated lattice of approximations
$\partial X(x)$ for every point $x \in X$. The semantics of an expression $e$ of type $\sigma$ with a single
free variable of type $\tau$ (with $\tau$ and $\sigma$ denoting $X$ and $Y$) must then also have two
components: a function $f: X \to Y$ giving the conventional interpretation of $e$, and a family of Galois
connections $\partial f(x): \partial X(x) \to \partial Y(f(x))$ for every $x \in X$ which can be used for
slicing over that expression.

This pattern of objects and morphisms is captured by the construction used in CHAD~\cite{vákár22,nunes2023} to
describe automatic differentiation, namely the \emph{Grothendieck construction} for a particular indexed
category $T: \Set^\op \to \Cat$. For CHAD $T(X)$ will be the functor category $\Func{X}{\FinVect}$, regarded
as the category of $X$-indexed families of finite vector spaces, with indexed linear maps; here $T(X) =
\Func{X}{\LatGal}$, the category of $X$-indexed families of bounded lattices, with indexed Galois connections
as morphisms. The Grothendieck construction $\Grothendieck{}T$~(\defref{Grothendieck} below) for an indexed
category $T$, sometimes called the \emph{total category} for $T$, is the category of all $\Set$-indexed
families of objects of $T$, together with morphisms that account for both structure within each family and how
families transform under reindexing.

@@

\subsection{Category of Families}
\label{sec:Fam}

@@

Now consider the total category $\Fam(\LatGal)$, which is our basic setting for interpreting first-order
programs for \GPS. This has as objects $(X, \partial X)$, all pairs of a set $X$ and and for every $x \in X$,
a bounded lattice $\partial X(x)$, and as morphisms $(X, \partial X) \to (Y, \partial Y)$, all pairs $(f,
\partial f)$ of a function $f: X \to Y$ and for every $x \in X$, a Galois connection $\partial f(x): \partial
X(x) \to \partial Y(f(x))$, as we sketched informally at the beginning of this section. The pullback
(reindexing) along $f$ in the composition $(g, \partial g) \comp (f, \partial f) = (g \comp f,
\reindex{\partial g}{f} \comp \partial f)$ selects the appropriate lattice of approximations and Galois
connection at each point, using the baseline (unapproximated) function $f$.

>>>>

The goal now is to equip the interpretation of a (first-order) program with a Galois connection relating
approximations of its inputs to approximations of its output. Since the specific lattices of approximations
depend on the value being approximated, the semantics of every type $\tau$ must have both a ``conventional''
component, providing the set $X$ which interprets $\tau$, plus an associated lattice of approximations
$\partial X(x)$ for every point $x \in X$. The semantics of an expression $e$ of type $\sigma$ with a single
free variable of type $\tau$ (with $\tau$ and $\sigma$ denoting $X$ and $Y$) must then also have two
components: a function $f: X \to Y$ giving the conventional interpretation of $e$, plus a family of Galois
connections $\partial f(x): \partial X(x) \to \partial Y(f(x))$ for every $x \in X$ which can be used for
slicing over that expression.

This pattern of objects and morphisms is captured by the construction used in CHAD~\cite{vákár22,nunes2023} to
describe automatic differentiation, namely the \emph{Grothendieck construction} for a particular indexed
category $T: \Set^\op \to \Cat$. For CHAD, $T(X)$ will be the functor category $\Func{X}{\FinVect}$, regarded
as the category of $X$-indexed families of finite vector spaces, with indexed linear maps; for Galois slicing
we will take $T(X) = \Func{X}{\LatGal}$, the category of $X$-indexed families of bounded lattices, with
indexed Galois connections as morphisms. The Grothendieck construction
$\Grothendieck{}T$~(\defref{Grothendieck} below) for an indexed category $T$, sometimes called the \emph{total
category} for $T$, is the category of all $\Set$-indexed families of objects of $T$, together with morphisms
that account for both structure within each family and how families transform under reindexing.

@@

\subsubsection{Category of Families}
\label{sec:Fam}

@@

Now consider the total category $\Fam(\LatGal)$, which is our basic setting for interpreting first-order
programs for \GPS. This has as objects $(X, \partial X)$, all pairs of a set $X$ and and for every $x \in X$,
a bounded lattice $\partial X(x)$, and as morphisms $(X, \partial X) \to (Y, \partial Y)$, all pairs $(f,
\partial f)$ of a function $f: X \to Y$ and for every $x \in X$, a Galois connection $\partial f(x): \partial
X(x) \to \partial Y(f(x))$. This is precisely the setup we identified informally at the beginning of this
section. The pullback (reindexing) along $f$ in the composition $(g, \partial g) \comp (f, \partial f) = (g
\comp f, \reindex{\partial g}{f} \comp \partial f)$ selects the appropriate lattice of approximations and
Galois connection at each point, using the baseline (unapproximated) function $f$.

\subsection{Automatic Differentiation via the Grothendieck Construction}

Automatic differentiation for first-order programs can also be recovered in this setup, via the total category
$\Fam(\FinVect)$.
